Hierarchy Builder: algebraic~hierarchies made easy in Coq with Elpi

### 2020-04-27 hb.tex
fix overfull line
overfull hbox code + making addrC proofs look like each other
comment sentence about flat PC performance
acknowledgments

  Definition #$b_{n+1}$# : #$f_{n+1}$# T .. := ...
  #\hbinstance{}# T #$b_{n+1}$#.
  ..
  Definition #$b_{n+k}$# : #$f_{n+k}$# T .. := ...
  #\hbinstance{}# T #$b_{n+k}$#.
#\hbend{}#.
\end{coqcode}
  Declares the record \coq{axioms} inside a module \coq{F}.
  This record \coq{F.axioms} depends on the mixins
  \(\requires{}(\coq{F.axioms}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\) and is registered
  as a factory.  It exports an
  \phantterm{} \coq{F.Build} and a notation
  \coq{F} standing for \coq{F.axioms}, so that the module name can be used to
  denote the \coq{axioms} record it contains.
  Finally it uses the \factoryinstances{}
  \(b_{n+1}, \ldots, b_{n+k}\) provided by the user in order to derive
  \mixinbuilders{}, so that
  \(\provides{}(\coq{F}) = \provides{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right).\)
  It is thus necessary that
  \(\requires{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right) \subseteq
  \provides{}^\star\left(\enum{f_1}{f_n}\right).\)
\end{commands}

@@

Next we define Abelian groups.
Since the monoid structure is the bottom of this hierarchy its class record
\coq{Monoid} consists of just one \mixin{}. In contrast the class record
of Abelian groups consists of two \mixins{} where the second one depends
 on the former (since Abelian groups inherit from monoids).
\begin{coqcode}
Module AbelianGroup.
Record axioms (T : Type) : Type := Class {
  Monoid_of_Type_mixin : Monoid_of_Type T;
  AbelianGroup_of_Monoid_mixin : AbelianGroup_of_Monoid T Monoid_of_Type_mixin; }.
Structure type : Type := Pack { sort : Type; class : axioms sort }.
End AbelianGroup.
\end{coqcode}
In the flat variant of Packed Classes, a class record gathers mixins as its
fields directly as in above. In contrast, a class record of the non-flat variant
packs a class instance of one of the superclasses with remaining mixins, which
reduces the amount of code to implement inheritance significantly.
Nested class records and chains of their projections exponentially slow down the
conversion while the flat variant can mitigate the issue. Since we do not need
to care about the amount of code in automated generation, \hb{} uses the flat
variant of Packed Classes as its target language.

@@

@@

  Lemma addrC : commutative (add : A -> A -> A).
  Proof.
  have addKl (a b c : A) : a + b = a + c -> b = c.
    apply: can_inj (add a) (add (opp_f a)) _ _ _.
    by move=> x; rewrite addrA addNr_f add0r.
  have addKr (a b c : A) : b + a = c + a -> b = c.
    apply: can_inj (add ^~ a) (add ^~ (opp_f a)) _ _ _.
    by move=> x; rewrite /= -addrA addrN_f addr0.
  have innerC (a b : A) : a + b + (a + b) = a + a + (b + b).
    by rewrite -[a+b]mul1r_f -mulrDl_f mulrDr_f !mulrDl_f !mul1r_f.
  move=> x y; apply: addKl (x) _ _ _; apply: addKr (y) _ _ _.
  by rewrite -!addrA [in RHS]addrA innerC !addrA.
  Qed.

>>>>

  Definition #$b_{n+1}$# : #$f_{n+1}$# T .. := ...
  #\hbinstance{}# T #$b_{n+1}$#.
  ..
  Definition #$b_{n+k}$# : #$f_{n+k}$# T .. := ...
  #\hbinstance{}# T #$b_{n+k}$#.
#\hbend{}#.
\end{coqcode}
  Declares the record \coq{axioms} inside a module \coq{F}.
  This record \coq{F.axioms} depends on the mixins
  \(\requires{}(\coq{F.axioms}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\) and is registered
  as a factory.  It exports an
  \phantterm{} \coq{F.Build} and a notation
  \coq{F} standing for \coq{F.axioms}, so that the module name can be used to
  denote the \coq{axioms} record it contains.
  Finally it uses the \factoryinstances{}
  \(b_{n+1}, \ldots, b_{n+k}\) provided by the user in order to derive
  \mixinbuilders{}, so that
  \(\provides{}(\coq{F}) = \provides{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right).\)\\
  It is thus necessary that
  \(\requires{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right) \subseteq
  \provides{}^\star\left(\enum{f_1}{f_n}\right).\)
\end{commands}

@@

Next we define Abelian groups.
Since the monoid structure is the bottom of this hierarchy its class record
\coq{Monoid} consists of just one \mixin{}. In contrast the class record
of Abelian groups consists of two \mixins{} where the second one depends
 on the former (since Abelian groups inherit from monoids).
\begin{coqcode}
Module AbelianGroup.
Record axioms (T : Type) : Type := Class {
  Monoid_of_Type_mixin : Monoid_of_Type T;
  AbelianGroup_of_Monoid_mixin : AbelianGroup_of_Monoid T Monoid_of_Type_mixin; }.
Structure type : Type := Pack { sort : Type; class : axioms sort }.
End AbelianGroup.
\end{coqcode}
In the flat variant of Packed Classes, a class record gathers mixins as its
fields directly as in above. In contrast, a class record of the non-flat variant
packs a class instance of one of the superclasses with remaining mixins, which
reduces the amount of code to implement inheritance significantly.
%Nested class records and chains of their projections exponentially slow down the
%conversion while the flat variant can mitigate the issue. 
Since we do not need
to care about the amount of code in automated generation, \hb{} uses the flat
variant of Packed Classes as its target language.

@@

\section*{Acknowledgments}

We are particularly grateful to Georges Gonthier, Assia Mahboubi and
Florent Hivert for the many discussions we had that helped to shape
various parts of this work, as well as other developers of the
Mathematical Components library and members of Dagstuhl seminar
18341. We also thank many Coq developers and users who showed interest
and support in this work. We thank the anonymous reviewers of this
paper for their useful comments.

@@

  Lemma addrC : commutative (add : A -> A -> A).
  Proof.
  have innerC (a b : A) : a + b + (a + b) = a + a + (b + b).
    by rewrite -[a+b]mul1r_f -mulrDl_f mulrDr_f !mulrDl_f !mul1r_f.
  have addKl (a b c : A) : a + b = a + c -> b = c.
    apply: can_inj (add a) (add (opp_f a)) _ _ _.
    by move=> x; rewrite addrA addNr_f add0r.
  have addKr (a b c : A) : b + a = c + a -> b = c.
    apply: can_inj (add ^~ a) (add ^~ (opp_f a)) _ _ _.
    by move=> x; rewrite /= -addrA addrN_f addr0.
  move=> x y; apply: addKl (x) _ _ _; apply: addKr (y) _ _ _.
  by rewrite -!addrA [in RHS]addrA innerC !addrA.
  Qed.
